---
title: "1M17 Positive-Control Docking Analysis"
author: "Mehnaz Hossain Meem"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: cosmo
    fig_width: 8
    fig_height: 6
  pdf_document:
    toc: true
    fig_width: 8
    fig_height: 6
---

## Analysis aim

This analysis evaluates positive-control redocking of crystallographic
erlotinib (AQ4) into the 1M17 EGFR structure. Only the top-ranked AutoDock
Vina pose (`MODEL 1`) is analysed from each output. Docking distance is the
Euclidean distance between the centre of the docked ligand and the
crystallographic erlotinib centre.

Expected filenames follow this pattern:

```text
prep_1M17_10_200.pdbqt
prep_1M17_10_400.pdbqt
...
prep_1M17_100_10000.pdbqt
```

## Load packages

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("ggplot2", "dplyr", "knitr")

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install these packages before knitting: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(ggplot2)
library(dplyr)
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Analysis settings

```{r settings}
data_dir <- "/mnt/data/mehnaz/1M17_Crys_Docking"

expected_box_sizes <- c(
  10, 15, 20, 25, 30, 40, 50, 60, 70, 80, 90, 100
)

crystal_center <- c(
  x = 21.6975,
  y = 0.3030,
  z = 52.0925
)

file_pattern <- paste0(
  "^prep_1M17_(",
  paste(expected_box_sizes, collapse = "|"),
  ")_[0-9]+\\.pdbqt$"
)

box_colours <- c(
  "10" = "purple",
  "15" = "red",
  "20" = "blue",
  "25" = "darkgreen",
  "30" = "orange",
  "40" = "brown",
  "50" = "deeppink",
  "60" = "cyan4",
  "70" = "black",
  "80" = "goldenrod3",
  "90" = "steelblue",
  "100" = "magenta4"
)
```

## Crystallographic ligand centre

```{r crystal-centre-table}
crystal_center_table <- data.frame(
  coordinate = c("x", "y", "z"),
  value = unname(crystal_center)
)

kable(
  crystal_center_table,
  digits = 4,
  caption = "Crystallographic erlotinib centre used for distance calculation"
)
```

## Functions

### Read MODEL 1

```{r model-one-function}
read_model_1 <- function(pdbqt_file) {
  lines <- readLines(pdbqt_file, warn = FALSE)

  model_start <- grep(
    "^MODEL[[:space:]]+1[[:space:]]*$",
    lines
  )[1]

  if (is.na(model_start)) {
    return(lines)
  }

  model_end <- which(
    seq_along(lines) > model_start & grepl("^ENDMDL", lines)
  )[1]

  if (is.na(model_end)) {
    stop("MODEL 1 has no ENDMDL in: ", pdbqt_file)
  }

  lines[model_start:model_end]
}
```

### Calculate the docked-ligand centre

The ligand centre is calculated using the midpoint between the minimum and
maximum coordinate on each axis, matching the method used to calculate the
crystallographic erlotinib centre.

```{r ligand-centre-function}
calculate_ligand_center <- function(model_lines) {
  atom_lines <- model_lines[
    grepl("^(ATOM|HETATM)", model_lines)
  ]

  if (length(atom_lines) == 0) {
    return(c(x = NA_real_, y = NA_real_, z = NA_real_))
  }

  x <- suppressWarnings(as.numeric(substr(atom_lines, 31, 38)))
  y <- suppressWarnings(as.numeric(substr(atom_lines, 39, 46)))
  z <- suppressWarnings(as.numeric(substr(atom_lines, 47, 54)))

  valid <- is.finite(x) & is.finite(y) & is.finite(z)
  x <- x[valid]
  y <- y[valid]
  z <- z[valid]

  if (length(x) == 0) {
    return(c(x = NA_real_, y = NA_real_, z = NA_real_))
  }

  c(
    x = mean(range(x)),
    y = mean(range(y)),
    z = mean(range(z))
  )
}
```

### Extract the MODEL 1 Vina score

```{r vina-score-function}
extract_vina_score <- function(model_lines, pdbqt_file) {
  score_line <- grep(
    "^REMARK VINA RESULT:",
    model_lines,
    value = TRUE
  )

  if (length(score_line) == 0) {
    all_lines <- readLines(pdbqt_file, warn = FALSE)
    score_line <- grep(
      "^REMARK VINA RESULT:",
      all_lines,
      value = TRUE
    )
  }

  if (length(score_line) == 0) {
    return(NA_real_)
  }

  suppressWarnings(
    as.numeric(
      sub(
        "^REMARK VINA RESULT:[[:space:]]*([-+]?[0-9]*\\.?[0-9]+).*$",
        "\\1",
        score_line[1]
      )
    )
  )
}
```

### Analyse one docking output

```{r analyse-one-file}
analyse_docking_file <- function(pdbqt_file) {
  filename <- basename(pdbqt_file)

  filename_match <- regexec(
    "^prep_1M17_([0-9]+)_([0-9]+)\\.pdbqt$",
    filename
  )

  values <- regmatches(filename, filename_match)[[1]]

  if (length(values) != 3) {
    stop("Unexpected docking filename: ", filename)
  }

  model_lines <- read_model_1(pdbqt_file)
  ligand_center <- calculate_ligand_center(model_lines)
  docking_score <- extract_vina_score(model_lines, pdbqt_file)

  if (any(!is.finite(ligand_center))) {
    distance_A <- NA_real_
  } else {
    distance_A <- sqrt(sum((ligand_center - crystal_center)^2))
  }

  data.frame(
    file = filename,
    box_size = as.numeric(values[2]),
    seed = as.numeric(values[3]),
    docking_score = docking_score,
    distance_A = distance_A,
    ligand_center_x = unname(ligand_center["x"]),
    ligand_center_y = unname(ligand_center["y"]),
    ligand_center_z = unname(ligand_center["z"]),
    stringsAsFactors = FALSE
  )
}
```

## Find docking files

```{r find-files}
docking_files <- list.files(
  path = data_dir,
  pattern = file_pattern,
  full.names = TRUE
)

if (length(docking_files) == 0) {
  stop(
    "No matching 1M17 docking files were found in: ",
    data_dir
  )
}

cat("Number of docking files found:", length(docking_files), "\n")
```

## Check seed-run counts

```{r file-counts}
file_information <- data.frame(
  file = basename(docking_files),
  stringsAsFactors = FALSE
)

file_information$box_size <- as.numeric(
  sub(
    "^prep_1M17_([0-9]+)_[0-9]+\\.pdbqt$",
    "\\1",
    file_information$file
  )
)

file_information$seed <- as.numeric(
  sub(
    "^prep_1M17_[0-9]+_([0-9]+)\\.pdbqt$",
    "\\1",
    file_information$file
  )
)

file_counts <- file_information |>
  group_by(box_size) |>
  summarise(
    number_of_files = n(),
    minimum_seed = min(seed),
    maximum_seed = max(seed),
    .groups = "drop"
  ) |>
  arrange(box_size)

kable(
  file_counts,
  caption = "Number of 1M17 docking files found for each box size"
)
```

## Analyse all outputs

```{r analyse-all-files}
results <- bind_rows(
  lapply(docking_files, analyse_docking_file)
) |>
  arrange(box_size, seed)

failed_rows <- results |>
  filter(!is.finite(docking_score) | !is.finite(distance_A))

valid_results <- results |>
  filter(is.finite(docking_score), is.finite(distance_A)) |>
  mutate(
    box = factor(
      box_size,
      levels = expected_box_sizes
    )
  )

if (nrow(valid_results) == 0) {
  stop("No valid docking scores and distances were extracted.")
}

cat("Valid docking files analysed:", nrow(valid_results), "\n")
cat("Files with missing values:", nrow(failed_rows), "\n")
```

## Files that could not be parsed

```{r failed-files}
if (nrow(failed_rows) == 0) {
  cat("All docking files were parsed successfully.")
} else {
  kable(
    failed_rows,
    digits = 4,
    caption = "Files with missing docking-score or coordinate data"
  )
}
```

## Summary by grid-box size

```{r summary-by-box}
box_summary <- valid_results |>
  group_by(box_size) |>
  summarise(
    number_of_runs = n(),
    mean_score = mean(docking_score),
    median_score = median(docking_score),
    best_score = min(docking_score),
    mean_distance_A = mean(distance_A),
    median_distance_A = median(distance_A),
    closest_distance_A = min(distance_A),
    .groups = "drop"
  ) |>
  arrange(box_size)

kable(
  box_summary,
  digits = 4,
  caption = "1M17 positive-control summary by grid-box size"
)
```

## Best-affinity MODEL 1 result from each box

The most-negative Vina score is selected from each box size.

```{r best-affinity-results}
best_affinity_results <- valid_results |>
  group_by(box_size) |>
  slice_min(
    order_by = docking_score,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup() |>
  arrange(box_size)

kable(
  best_affinity_results |>
    select(box_size, seed, docking_score, distance_A, file),
  digits = 5,
  caption = "Best-affinity MODEL 1 result from each 1M17 grid box"
)
```

## Closest MODEL 1 result from each box

```{r closest-results}
closest_results <- valid_results |>
  group_by(box_size) |>
  slice_min(
    order_by = distance_A,
    n = 1,
    with_ties = FALSE
  ) |>
  ungroup() |>
  arrange(box_size)

kable(
  closest_results |>
    select(box_size, seed, distance_A, docking_score, file),
  digits = 5,
  caption = "Closest MODEL 1 erlotinib pose from each 1M17 grid box"
)
```

## Boxplot: docking distance by grid-box size

```{r distance-boxplot, fig.width=9, fig.height=6}
distance_boxplot <- ggplot(
  valid_results,
  aes(x = box, y = distance_A)
) +
  geom_boxplot(
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  geom_jitter(
    width = 0.18,
    alpha = 0.50,
    size = 1.4,
    colour = "darkgray"
  ) +
  geom_point(
    data = best_affinity_results,
    aes(
      x = factor(box_size, levels = expected_box_sizes),
      y = distance_A,
      fill = factor(box_size)
    ),
    inherit.aes = FALSE,
    shape = 21,
    size = 4,
    stroke = 0.8,
    colour = "black"
  ) +
  scale_fill_manual(values = box_colours, guide = "none") +
  labs(
    title = "1M17 Docking Distance by Grid Box Size",
    subtitle = "Positive control; exhaustiveness = 12",
    x = "Grid box size (Å)",
    y = "Docking distance (Å)"
  ) +
  theme_classic(base_size = 12)

distance_boxplot
```

## Boxplot: Vina score by grid-box size

```{r score-boxplot, fig.width=9, fig.height=6}
score_boxplot <- ggplot(
  valid_results,
  aes(x = box, y = docking_score)
) +
  geom_boxplot(
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  geom_jitter(
    width = 0.18,
    alpha = 0.50,
    size = 1.4,
    colour = "darkgray"
  ) +
  labs(
    title = "1M17 Vina Scores by Grid Box Size",
    subtitle = "Positive control; exhaustiveness = 12",
    x = "Grid box size (Å)",
    y = "Vina docking score (kcal/mol)"
  ) +
  theme_classic(base_size = 12)

score_boxplot
```

## Scatterplot: best affinity versus distance

Each point is the most-negative `MODEL 1` Vina score selected from one box.

```{r scatter-plot, fig.width=9, fig.height=6}
scatter_df <- best_affinity_results |>
  mutate(
    box = factor(
      box_size,
      levels = expected_box_sizes
    )
  )

scatter_plot <- ggplot(
  scatter_df,
  aes(
    x = distance_A,
    y = docking_score,
    colour = box,
    label = paste0(box_size, " Å")
  )
) +
  geom_point(size = 4) +
  geom_text(
    nudge_y = 0.08,
    show.legend = FALSE,
    check_overlap = TRUE
  ) +
  scale_colour_manual(values = box_colours, drop = TRUE) +
  labs(
    title = "1M17 Best Binding Affinity for Each Grid Box",
    subtitle = "Positive-control erlotinib redocking",
    x = "Docking distance (Å)",
    y = "Best Vina binding affinity (kcal/mol)",
    colour = "Grid box size"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "right")

scatter_plot
```

## Save results and plots

```{r save-results}
write.csv(
  valid_results,
  file.path(data_dir, "1M17_positive_control_results.csv"),
  row.names = FALSE
)

write.csv(
  box_summary,
  file.path(data_dir, "1M17_positive_control_box_summary.csv"),
  row.names = FALSE
)

write.csv(
  best_affinity_results,
  file.path(data_dir, "1M17_positive_control_best_affinity.csv"),
  row.names = FALSE
)

write.csv(
  closest_results,
  file.path(data_dir, "1M17_positive_control_closest_poses.csv"),
  row.names = FALSE
)

ggsave(
  file.path(data_dir, "1M17_positive_control_distance_boxplot.png"),
  distance_boxplot,
  width = 9,
  height = 6,
  dpi = 300
)

ggsave(
  file.path(data_dir, "1M17_positive_control_score_boxplot.png"),
  score_boxplot,
  width = 9,
  height = 6,
  dpi = 300
)

ggsave(
  file.path(data_dir, "1M17_positive_control_score_vs_distance.png"),
  scatter_plot,
  width = 9,
  height = 6,
  dpi = 300
)
```

## Interpretation guide

- A smaller docking distance indicates that the docked erlotinib pose is
  closer to the crystallographic ligand location.
- A more-negative Vina score indicates stronger predicted binding.
- The best-scoring pose and closest pose can come from different seed runs.
- Results from 10 and 15 Å boxes should be interpreted cautiously because
  erlotinib spans approximately 15.7 Å along its longest axis.

## Session information

```{r session-info}
sessionInfo()
```
