---
title: "Calculate distance between AutoDock Vina ligand distance and crystal structure"
author: "Mehnaz Meem"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: TBA

## Introduction

This analysis calculates the distance between the centre of the top-ranked
AutoDock Vina ligand pose and the crystal-ligand centre for 3ERT. It also
compares docking distance with Vina predicted binding affinity.

## Crystal structure position

```{r crystal}
cx <- 30.282
cy <- -1.913
cz <- 24.2065

cxyz <- c(cx, cy, cz)
```

## Read example data

```{r autodock_data}
pdb <- readLines(
  "~/Protein_Comparison_workflows/prep_3ERT_10_200.pdbqt",
  warn = FALSE
)

# Keep only the top-ranked Vina pose (MODEL 1)
model_start <- grep("^MODEL[[:space:]]+1[[:space:]]*$", pdb)[1]

if (!is.na(model_start)) {
  model_end <- which(
    seq_along(pdb) > model_start & grepl("^ENDMDL", pdb)
  )[1]

  if (is.na(model_end)) {
    stop("MODEL 1 has no ENDMDL in the example PDBQT file")
  }

  pdb <- pdb[model_start:model_end]
}

atoms <- pdb[grep("^(ATOM|HETATM)", pdb)]

if (length(atoms) == 0) {
  stop("No ATOM or HETATM lines found in the example PDBQT file")
}
```

Extract the x, y and z coordinates from their fixed PDBQT positions.

```{r extract_coords}
x <- as.numeric(substr(atoms, 31, 38))
y <- as.numeric(substr(atoms, 39, 46))
z <- as.numeric(substr(atoms, 47, 54))

if (anyNA(c(x, y, z))) {
  stop("Could not read coordinates from the example PDBQT file")
}

mx <- cbind(x = x, y = y, z = z)

str(mx)
summary(mx)
```

## Predicted ligand centre and distance

```{r example_distance}
px <- mean(range(mx[, "x"]))
py <- mean(range(mx[, "y"]))
pz <- mean(range(mx[, "z"]))

dx <- abs(cx - px)
dy <- abs(cy - py)
dz <- abs(cz - pz)

hyp <- sqrt(dx^2 + dy^2 + dz^2)
hyp
```

## Define distance function

```{r dist_function}
model_ligand_dist <- function(cxyz, pdbqt_file) {
  pdb <- readLines(pdbqt_file, warn = FALSE)

  # Keep only the top-ranked Vina pose (MODEL 1)
  model_start <- grep("^MODEL[[:space:]]+1[[:space:]]*$", pdb)[1]

  if (!is.na(model_start)) {
    model_end <- which(
      seq_along(pdb) > model_start & grepl("^ENDMDL", pdb)
    )[1]

    if (is.na(model_end)) {
      stop(paste("MODEL 1 has no ENDMDL in:", pdbqt_file))
    }

    pdb <- pdb[model_start:model_end]
  }

  atoms <- pdb[grep("^(ATOM|HETATM)", pdb)]

  if (length(atoms) == 0) {
    stop(paste("No ATOM or HETATM lines found in:", pdbqt_file))
  }

  # Read coordinates from fixed-width PDBQT columns
  x <- as.numeric(substr(atoms, 31, 38))
  y <- as.numeric(substr(atoms, 39, 46))
  z <- as.numeric(substr(atoms, 47, 54))

  if (anyNA(c(x, y, z))) {
    stop(paste("Could not read coordinates from:", pdbqt_file))
  }

  predicted_centre <- c(
    mean(range(x)),
    mean(range(y)),
    mean(range(z))
  )

  sqrt(sum((cxyz - predicted_centre)^2))
}
```

Test the function.

```{r functest}
model_ligand_dist(
  cxyz,
  "~/Protein_Comparison_workflows/prep_3ERT_10_200.pdbqt"
)
```

## Run for all PDBQT files

```{r runall}
workdir <- "~/Protein_Comparison_workflows"
box_pattern <- "10|15|20|25|30|40|50|60|70|80|90|100"

myfiles <- list.files(
  path = workdir,
  pattern = paste0(
    "^prep_3ERT_(",
    box_pattern,
    ")_[0-9]+\\.pdbqt$"
  ),
  full.names = TRUE
)

if (length(myfiles) == 0) {
  stop("No 3ERT docking files found in ~/Protein_Comparison_workflows")
}

distance <- vapply(
  myfiles,
  function(f) model_ligand_dist(cxyz, f),
  numeric(1)
)

plot_df <- data.frame(
  file = basename(myfiles),
  distance = as.numeric(distance),
  stringsAsFactors = FALSE
)

plot_df$box <- sub(
  paste0(
    "^prep_3ERT_(",
    box_pattern,
    ")_[0-9]+\\.pdbqt$"
  ),
  "\\1",
  plot_df$file
)

plot_df$seed <- as.numeric(
  sub(
    paste0(
      "^prep_3ERT_(?:",
      box_pattern,
      ")_([0-9]+)\\.pdbqt$"
    ),
    "\\1",
    plot_df$file,
    perl = TRUE
  )
)

plot_df$box <- factor(
  plot_df$box,
  levels = c(
    "10", "15", "20", "25", "30", "40", "50",
    "60", "70", "80", "90", "100"
  )
)

plot_df <- plot_df[order(plot_df$box, plot_df$seed), ]
```

## Extract Vina predicted binding affinity

```{r vina_energy}
extract_vina_energy <- function(pdbqt_file) {
  lines <- readLines(pdbqt_file, warn = FALSE)

  vina_line <- grep(
    "^REMARK VINA RESULT:",
    lines,
    value = TRUE
  )

  if (length(vina_line) == 0) {
    return(NA_real_)
  }

  as.numeric(
    sub(
      "^REMARK VINA RESULT:[[:space:]]*([-0-9.]+).*",
      "\\1",
      vina_line[1]
    )
  )
}

energy_named <- vapply(
  myfiles,
  extract_vina_energy,
  numeric(1)
)

names(energy_named) <- basename(myfiles)
plot_df$free_energy <- as.numeric(energy_named[plot_df$file])

print(
  plot_df[, c("file", "box", "seed", "distance", "free_energy")]
)
```

## Summary statistics

```{r summary_statistics}
summary_stats <- aggregate(
  distance ~ box,
  data = plot_df,
  function(x) c(
    Mean = mean(x),
    Median = median(x),
    SD = sd(x),
    Min = min(x),
    Max = max(x)
  )
)

print(summary_stats)
```

##Best binding affinity result from each box

```{r closest_results}
best_affinity_results <- do.call(
  rbind,
  lapply(
    split(plot_df, plot_df$box, drop = TRUE),
    function(df) {

      # Remove files where Vina affinity was not extracted
      df <- df[!is.na(df$free_energy), ]

      if (nrow(df) == 0) {
        return(NULL)
      }

      # More negative affinity represents stronger predicted binding
      df[which.min(df$free_energy), ]
    }
  )
)

if (is.null(best_affinity_results) ||
    nrow(best_affinity_results) == 0) {
  stop("No best-affinity results could be selected")
}

best_affinity_results <- best_affinity_results[
  order(
    as.numeric(
      as.character(best_affinity_results$box)
    )
  ),
]

rownames(best_affinity_results) <- NULL

print(
  best_affinity_results[, c(
    "box",
    "seed",
    "free_energy",
    "distance",
    "file"
  )]
)
```

## Docking distance by grid box

```{r beeswarm_boxplot, fig.width=8, fig.height=6}
if (!requireNamespace("beeswarm", quietly = TRUE)) {
  stop("Install beeswarm first using install.packages('beeswarm')")
}

dist_list <- split(
  plot_df$distance,
  plot_df$box,
  drop = TRUE
)

boxplot(
  dist_list,
  xlab = "Grid box size (Å)",
  ylab = "Docking distance (Å)",
  main = "3ERT Docking Distance by Grid Box Size\nExhaustiveness = 12",
  col = "white",
  border = "black"
)

beeswarm::beeswarm(
  dist_list,
  add = TRUE,
  pch = 19,
  cex = 0.8,
  col = "darkgray"
)
```

## Vina binding affinity versus distance

```{r best_affinity_vs_distance, fig.width=8, fig.height=6}

scatter_df <- best_affinity_results[
  complete.cases(
    best_affinity_results$distance,
    best_affinity_results$free_energy
  ),
]

if (nrow(scatter_df) == 0) {
  stop(
    paste(
      "No complete best-affinity and distance",
      "values were found"
    )
  )
}

# Colours assigned to each grid-box size
box_colours <- c(
  "10" = "purple",
  "15" = "red",
  "20" = "blue",
  "25" = "darkgreen",
  "30" = "orange",
  "40" = "brown",
  "50" = "deeppink",
  "60" = "cyan4",
  "70" = "black",
  "80" = "goldenrod3",
  "90" = "steelblue",
  "100" = "magenta4"
)

point_colours <- box_colours[
  as.character(scatter_df$box)
]

plot(
  scatter_df$distance,
  scatter_df$free_energy,
  pch = 19,
  cex = 1.8,
  col = point_colours,
  xlab = "Docking distance (Å)",
  ylab = "Best Vina binding affinity (kcal/mol)",
  main = paste(
    "3ERT Best Binding Affinity",
    "for Each Grid Box"
  )
)

# Add box-size labels beside each point
text(
  scatter_df$distance,
  scatter_df$free_energy,
  labels = paste0(
    scatter_df$box,
    " Å"
  ),
  pos = 3,
  cex = 0.8,
  col = point_colours
)

legend(
  "topright",
  legend = paste0(
    levels(droplevels(scatter_df$box)),
    " Å"
  ),
  col = box_colours[
    levels(droplevels(scatter_df$box))
  ],
  pch = 19,
  pt.cex = 1.3,
  title = "Grid box size",
  bty = "n"
)
```


## Session information

```{r session}
sessionInfo()
```
